Causal graph analysis of COVID-19 observational data in German districts reveals effects of determining factors on new reported cases

### 2021-03-19 manuscript/v4_medrxiv_revision/manuscript.tex
update table adj sets

We used a directed acyclic graph (DAG) \citep{schipf_directed_2011, textor_robust_2017} as a tool to analyze the causal relationships between several exposures and SARS-CoV-2 spread. To get an overview on published associations, a scoping review was conducted from 20th to 22nd of May 2020 within Pubmed and Google scholar. Restrictions were applied to English and German language and the publication date in the last one year. The following search terms were applied to abstracts and title in Pubmed (``COVID-19'' OR ``COVID19'' OR ``Corona'' OR ``Coronavirus'' OR ``SARS-CoV-2'') and connected separately in each case with a selection of exposure variables (``mobility'', ``public awareness'', ``awareness'', ``google trends'', ``ambient temperature'', ``temperature''). For ``mobility'', we analyzed \(n=8\) studies, \(N=103\) were scanned in Pubmed, together with the first ten pages (100 results) in Google Scholar (``awareness''/``public awareness''/``google trends'' \(n=9\), \(N=215\); ``temperature''/``ambient temperature'' \(n=16\), \(N=235\)). We integrated these findings where possible into the construction of our DAG, which can be seen in Figure \ref{fig:dag-covid-19}.

@@

We collected and aggregated data on reported COVID-19 cases, regional socio-demographic factors, weather, and general mobility on district and state level in Germany for the period of 15 February 2020 to 8 July 2020. Our observation period for the outcome consisted of all dates from 28 February 2020 to 8 July 2020 (\(T=132\)), since we used a lag of 12 days for all confounders. We did not exclude any states or districts (\(K=401\)). We analyzed the daily reported number of new cases as outcome (\(K\cdot T=54{\ }937\) observations). The set of possible predictors was derived from our causal DAG (see Table \ref{tab:model-vars} and Figure \ref{fig:dag-covid-19}). Due to modelling and data limitations, some of the predictors were unobserved or were modelled as a construct consisting of several variables. For our causal graph analysis of effects, we computed adjustment sets separately for all observed exposures within the DAG (if the respective exposure was identifiable within the DAG causal analysis framework).

@@

To assess the mobility of the German population, we used data publicly available on German state level from Google \citep{google_mobility}. Measurements are daily relative changes of mobility in percent compared to the period of 3 January 2020 to 6 February 2020. Missing values (\(25\) out of \(13{\ }488\)) were imputed with value \(0\) and the state level measurements were passed onto districts within the corresponding state. Google mobility data was available for six different sectors of daily life (``retail and recreation'', ``grocery and pharmacy'', ``parks'', ``transit stations'', ``workplaces'', ``residential'') which means that ``mobility'' is a construct consisting of several variables. All variables but ``residential'' mobility are relative changes of daily visitor numbers to the corresponding sectors compared to the reference period. ``Residential'' mobility is the relative change of daily time spent at residential areas. The six mobility variables showed high correlations among each other and with other variables. To reduce multicollinearity, we transformed them by principal component analysis (PCA) onto six uncorrelated principal components which were used in place of the original variables.

@@

\caption{\label{tab:model-vars}Observed model variables}
\centering
\fontsize{6}{8}\selectfont
\begin{tabular}[t]{llll>{\raggedright\arraybackslash}p{10em}l}
\toprule
variable & dynamics & level & type & unit/comment & source\\
\midrule
Weekday & daily & national & categorical & Sat through Thu as six binary variables, Fri as baseline & -\\
Holiday (report) & daily & national & binary & - & -\\
Holiday (exposure) & daily & national & binary & - & -\\
\addlinespace[0.3em]
\multicolumn{6}{l}{\textbf{Mobility}}\\
\hspace{1em}retail and recreation & daily & state & numeric & percent change compared to reference period & Google \citep{google_mobility}\\
\hspace{1em}grocery and pharmacy & daily & state & numeric & percent change compared to reference period & Google \citep{google_mobility}\\
\hspace{1em}parks & daily & state & numeric & percent change compared to reference period & Google \citep{google_mobility}\\
\hspace{1em}workplaces & daily & state & numeric & percent change compared to reference period & Google \citep{google_mobility}\\
\hspace{1em}residential & daily & state & numeric & percent change compared to reference period & Google \citep{google_mobility}\\
\hspace{1em}transit stations & daily & state & numeric & percent change compared to reference period & Google \citep{google_mobility}\\
\addlinespace[0.3em]
\multicolumn{6}{l}{\textbf{Weather}}\\
\hspace{1em}Rainfall & daily & district & numeric & mm (l/sqm) & DWD \citep{dwd_weather}\\
\hspace{1em}Temperature & daily & district & numeric & °C & DWD \citep{dwd_weather}\\
\hspace{1em}Humidity & daily & district & numeric & relative humidity (\%) & DWD \citep{dwd_weather}\\
\hspace{1em}Wind & daily & district & numeric & m/s & DWD \citep{dwd_weather}\\
\addlinespace[0.3em]
\multicolumn{6}{l}{\textbf{Policies}}\\
\hspace{1em}Ban of mass gatherings & daily & national & binary & - & -\\
\hspace{1em}School and kindergarten closures & daily & state & numeric & 0 for no closure, 1 for full closure, 0.5 for partial reopening & -\\
\hspace{1em}Contact restrictions & daily & national & binary & - & -\\
\hspace{1em}Mandatory face masks & daily & district & binary & - & IZA \citep{mitze2020face}\\
\addlinespace[0.3em]
\multicolumn{6}{l}{\textbf{Socio-demographic}}\\
\hspace{1em}Age & constant & district & numeric & 2 variables: share of population >=65 years \& <18 years & INKAR \citep{inkar}\\
\hspace{1em}Gender & constant & district & numeric & share of female population & INKAR \citep{inkar}\\
\hspace{1em}Population density & constant & district & numeric & population per sqkm & INKAR \citep{inkar}\\
\hspace{1em}Foreign citizens & constant & district & numeric & 2 variables: share of foreign citizens \& of population seeking refuge & INKAR \citep{inkar}\\
\hspace{1em}Socio-economic status & constant & district & numeric & share of households with low income & INKAR \citep{inkar}\\
\hspace{1em}Turnout & constant & district & numeric & voter turnout in last election & INKAR \citep{inkar}\\
\hspace{1em}Right-wing populist party votes & constant & district & numeric & share of votes for AfD in last election & INKAR \citep{inkar}\\
\hspace{1em}Nursing homes & constant & district & numeric & number of nursing (retirement) homes & INKAR \citep{inkar}\\
\addlinespace[0.3em]
\multicolumn{6}{l}{\textbf{Awareness}}\\
\hspace{1em}Searches corona & daily & state & numeric & percent relative to other states and observation period & Google \citep{google_trends}\\
\hspace{1em}COVID-19 burden & daily & district & numeric & reported cases on day of exposure & RKI \citep{casenumbers_rki}\\
\addlinespace[0.3em]
\multicolumn{6}{l}{\textbf{Case numbers}}\\
\hspace{1em}Reported new cases of COVID-19 & daily & district & numeric & - & RKI \citep{casenumbers_rki}\\
\hspace{1em}Active cases & daily & district & numeric & active cases on day of report & RKI \citep{casenumbers_rki}\\
\bottomrule
\end{tabular}
\end{table}

All variables but the outcome ``Reported new cases of COVID-19'' and the offset ``Active cases'' were centered for numerical stability. We scaled continuous variables by dividing by 2 times their standard deviation, while we left binary variables as-is \citep{gelman2008scaling}. After estimating their effects, we re-scaled variables to their original scale and for mobility, we re-transformed the effects of the principal components to the original mobility variables. Additionally, we lagged the effect of all variables (but outcome, offset, and the non-dynamic socio-demographic variables) by 12 days (see Section \ref{results}) which means that we assumed that their effects on the outcome will be visible after 12 days.

@@

We used a directed acyclic graph as a graphical representation of the hypothesized causal reasoning that leads to exposure to the SARS-CoV-2 virus, onset of COVID-19, and finally reports of COVID-19 cases. We use the terms ``causal effect'' or ``causal relationship'' for effect estimates that are based on this causal graph framework. Every node \(v_i\) in the causal graph is the graphical representation of an observed or unobserved variable \(x_i\), a directed edge \(e_{ij}\) is an arrow from node \(v_i\) to \(v_j\) that implies a direct causal relationship from variable \(x_i\) onto variable \(x_j\). The set of all nodes is denoted by \(V\), the set of all edges by \(E\), as such, the complete DAG is the tuple \(G=(V,E)\). The seminal works of Spirtes and Pearl \citep{spirtes2000causation, pearl2009causality} introduce the theory of causal analysis, do-calculus, and how to analyze a DAG to estimate the total or direct causal effect from a variable \(x_i\) onto a variable \(x_j\). The direct effect is the effect associated with the edge \(e_{ij}\) only (if it exists), while the total effect takes indirect effects via other paths from \(v_i\) to \(v_j\) into account, too. Here we estimated total effects only, since most of our variables were not hypothesized to have a direct effect on the \emph{reported} number of new COVID-19 cases. In contrast to prediction tasks, where one would include all variables available, it is actually ill-advised to use all available variables to estimate causal effects, due to introducing bias by adjusting for unnecessary variables within the causal DAG. This is why we need to identify a valid set of necessary variables (an adjustment set) to estimate the proper causal effect \citep{pearl2009causality}. The ``minimal adjustment set'' \citep{greenland1999causal} is a valid adjustment set of variables that does not contain another valid adjustment set as a subset. However, identifying a minimal adjustment set might not be enough to reliably estimate the causal effect. Thus, we identified the ``optimal adjustment set'' \citep{henckel2019graphical} as the set of variables which is a valid adjustment set while having the lowest asymptotic variance in the resulting causal effect estimates.

We analyzed the DAG from Section \ref{causal-model} with the R Software \citep{rsoftware} and the R packages \texttt{dagitty} (formal representation of the graph and minimal adjustment sets \citep{textor_robust_2017}) and \texttt{pcalg} (for finding an optimal adjustment set \citep{pcalg}). For the defined exposures and the outcome ``Reported new cases of COVID-19'', we computed the minimal and optimal adjustment sets. Since it was possible that these sets contained unobserved variables that needed to be left out of the regression model, we chose the valid set with the highest pseudo-\(R^2\) (see next section) to estimate the final total causal effect from exposure to outcome.

@@

To reduce the effects of multicollinearity and to prevent overfitting, we used a ridge regression approach \citep{ridge1970}, which is a regularization method that shrinks regression coefficients to allow for better fits on unseen data and alleviates the effect of correlated variables on their respective regression coefficients. The hyper-parameter \(\lambda\) of the ridge regression was chosen by 10-fold cross-validation, where the folds were constructed from subsets of the \(401\) districts.

We conducted a log-linear regression (function \texttt{glm} with \texttt{family=poisson()} for Poisson regression, and \texttt{glm.nb} from the \texttt{MASS} package for the negative binomial regression \citep{mass}) for the full data set to assess general model adequacy and to estimate the \(\theta\) parameter of the negative binomial. We used this parameter with the \texttt{cv.glmnet} function from the R package \texttt{glmnet} \citep{glmnet2010} with \texttt{family=negative.binomial(theta)} and chose the \(\lambda\) value within one standard deviation from the minimal \(\lambda\) as regularization hyper-parameter. Afterwards, we calculated the effects of separate exposures on the outcome. For every exposure, we analyzed the different valid adjustment sets given by analysis of the causal DAG (i.e.~the minimal and optimal adjustment sets). Then, we first checked if the respective set included unobserved variables. If this was the case for the optimal adjustment set, we discarded the unobserved variables from the set and checked if it was still a valid adjustment set (function \texttt{gac} in package \texttt{pcalg} \citep{perkovi2015complete}). If a minimal adjustment set contained unobserved variables, we discarded the whole set. If no valid adjustment set for a given exposure was available, we concluded that the effect of this exposure was unidentifiable within our causal graph. We used the function \texttt{glmnet} with the parameters \(\theta\) and \(\lambda\) as above on every remaining valid adjustment set as regressors and calculated the Akaike information criterion (AIC) for this set of regressors. Finally, for every exposure, we decided for the model/adjustment set (if available) with the lowest AIC. We report the exponentiated estimated coefficients.

@@

\caption{\label{tab:causaleffects}Causal effect estimates with 99 percent confidence bands}
\centering
\begin{tabular}[t]{lrrr}
\toprule
cause & effect estimate & CI (lower) & CI (upper)\\
\midrule
\addlinespace[0.3em]
\multicolumn{4}{l}{\textbf{Mobility}}\\
\hspace{1em}PC1 & 0.9044 & - & -\\
\hspace{1em}PC2 & 0.9823 & - & -\\
\hspace{1em}PC3 & 1.0465 & - & -\\
\hspace{1em}PC4 & 1.0247 & - & -\\
\hspace{1em}PC5 & 1.0037 & - & -\\
\hspace{1em}PC6 & 0.9999 & - & -\\
\addlinespace[0.3em]
\multicolumn{4}{l}{\textbf{Awareness}}\\
\hspace{1em}Searches corona & 1.0101 & 1.0085 & 1.0118\\
\hspace{1em}COVID-19 burden & 0.9005 & 0.8819 & 0.9202\\
\addlinespace[0.3em]
\multicolumn{4}{l}{\textbf{Weather}}\\
\hspace{1em}Temperature & 0.9460 & 0.9129 & 0.9803\\
\hspace{1em}Rainfall & 1.0108 & 1.0066 & 1.0150\\
\hspace{1em}Humidity & 1.0022 & 1.0006 & 1.0039\\
\hspace{1em}Wind & 0.9893 & 0.9802 & 0.9985\\
\bottomrule
\end{tabular}
\end{table}

We list the results of our causal analysis for the effects of our variables in Table \ref{tab:causaleffects}. The estimates are multiplicative rates of increase/decrease for a one unit increase of the respective variable: Values above 1 lead to an increase, below 1 to a decrease of the infection rate. To put these estimates into perspective, Figure \ref{fig:effectsrange} shows the relative causal effect of the different exposure variables on the number of reported COVID-19 cases on a range of sensible values of the exposure variables (95 percent quantiles of data points).

@@

Within our framework, we saw significant effects for mobility in retail/recreational areas and essential shopping (grocery and pharmacy). For retail/recreation, an increase of 1 percent point mobility compared to the reference period (03 January to 06 February 2020) leads to an increase of the daily reported case number by about percent. Contrarily, a corresponding increase of 1 percent point for the areas of grocery/pharmacy leads to a decrease in the reported case number by approximately percent. Mobility on workplaces showed a small effect of increase in case numbers for every 1 percent point increase in mobility. Other causal effects of mobility were insubstantial and not consistent in their direction (99 percent confidence intervals of estimates include 1). Figure \ref{fig:effectsrange} shows the effects of mobility on a range of possible values. Thus, we expect an increase of daily cases by approximately percent if mobility in retail/recreation reaches baseline levels of 0 percent difference to the reference period. On the other hand, an increase of mobility for grocery/pharmacy by 10 percent points compared to the reference period leads to a reduction of the infection rate by approximately percent.

``Awareness'' had two opposite effects on the outcome in our DAG. Awareness measured by Google searches for \emph{corona} had a positive effect on the number of reported cases. An one percent point increase of the state's Google searches (relative to other states and the observation period) leads to an increase of approximately 1 percent. For example, if a district shows 10 percent points more relative searches for \emph{corona} than another one, we expect approximately 11 percent more infections for this district after 8 days. \emph{COVID-19 burden} (reported number of cases on day of exposure) affected the outcome negatively, where every additional daily case in the district leads to a 9.9 percent decrease in newly reported case numbers. The last plot in Figure \ref{fig:effectsrange} visualizes this relationship: For a local outbreak with 20 daily cases as COVID-19 burden, we estimate as total causal effect a subsequent reduction of infection rate by 87.7 percent.

Within our model, we observed a causal effect of temperature and all other weather variables. Every increase of 1 degree Celsius in temperature leads to a reduction of the daily reported case numbers by approximately 5.4 percent. On the other hand, we found an increasing effect of rainfall: One millimeter (=1 liter per square meter) more rainfall leads to an increase of reported case numbers by approximately 1.1 percent. We observe effects for humidity and wind as well (higher humidity leading to more cases, stronger wind leading to less cases). In perspective (Figure \ref{fig:effectsrange}), with temperature we expect an increase by approximately 203.4 percent at a daily average temperature of \(0^{\circ}C\) compared to a day with \(20^{\circ}C\). For rainfall, we expect on a rainy day with 10 mm rainfall a corresponding increase of the infection rate by approximately 11.4 percent compared to a day with no precipitation.

In all cases we opted to use the reduced optimal adjustment set over the minimal adjustment sets because of higher pseudo-\(R^2\) values (mostly above 0.3), except for mobility, where the minimal adjustment set had a higher pseudo-\(R^2\). Notably, these sets always include most of our socio-demographic variables as confounders as well as the policy variables (cf.~Table \ref{tab:adjsets}, with the exception being COVID-19 burden). We also decided for the lag of 8 days based on the highest pseudo-\(R^2\) values compared to other lags on the chosen adjustment sets. Similarly, negative binomial regression was chosen over Poisson regression, because the latter showed overdispersion and overall lower pseudo-\(R^2\) values.

>>>>

We used a directed acyclic graph (DAG) \citep{schipf_directed_2011, textor_robust_2017} as a tool to analyze the causal relationships between several exposures and SARS-CoV-2 spread. To get an overview on published associations, a scoping review was conducted from 20th to 22nd of May 2020 within Pubmed and Google Scholar. Restrictions were applied to English and German language and the publication date in the last one year. The following search terms were applied to abstracts and titles in Pubmed (``COVID-19'' OR ``COVID19'' OR ``Corona'' OR ``Coronavirus'' OR ``SARS-CoV-2'') and connected separately in each case with a selection of exposure variables (``mobility'', ``public awareness'', ``awareness'', ``google trends'', ``ambient temperature'', ``temperature''). For ``mobility'', we analyzed \(n=8\) studies, \(N=103\) were scanned in Pubmed, together with the first ten pages (100 results) in Google Scholar (``awareness''/``public awareness''/``google trends'' \(n=9\), \(N=215\); ``temperature''/``ambient temperature'' \(n=16\), \(N=235\)). We integrated these findings where possible into the construction of our DAG, which can be seen in Figure \ref{fig:dag-covid-19}.

@@

We collected and aggregated data on reported COVID-19 cases, regional socio-demographic factors, weather, and general mobility on district and state level in Germany for the period of 15 February 2020 to 8 July 2020. Our observation period for the outcome consisted of all dates from 27 February 2020 to 8 July 2020 (\(T=133\)), since we used a lag of 12 days for all confounders. We did not exclude any states or districts (\(K=401\)). We analyzed the daily reported number of new cases as outcome (\(K\cdot T=53{\ }333\) observations). The set of possible predictors was derived from our causal DAG (see Table \ref{tab:model-vars} and Figure \ref{fig:dag-covid-19}). Due to modelling and data limitations, some of the predictors were unobserved or were modelled as a construct consisting of several variables. For our causal graph analysis of effects, we computed adjustment sets separately for all observed exposures within the DAG (if the respective exposure was identifiable within the DAG causal analysis framework).

@@

To assess the mobility of the German population, we used data publicly available on German state level from Google \citep{google_mobility}. Measurements are daily relative changes of mobility in percent compared to the period of 3 January 2020 to 6 February 2020. Missing values (\(25\) out of \(13{\ }488\)) were imputed with value \(0\) and the state level measurements were passed onto districts within the corresponding state. Google mobility data was available for six different sectors of daily life (``retail and recreation'', ``grocery and pharmacy'', ``parks'', ``transit stations'', ``workplaces'', ``residential'') which means that ``mobility'' is a construct consisting of several variables. All variables but ``residential'' mobility are relative changes of daily visitor numbers to the corresponding sectors compared to the reference period. ``Residential'' mobility is the relative change of daily time spent at residential areas. The six mobility variables showed high correlations among each other and with other variables. To reduce multicollinearity, we transformed them by principal component analysis (PCA) into six uncorrelated principal components which were used in place of the original variables.

@@

\caption{\label{tab:model-vars}Observed model variables}
\centering
\fontsize{6}{8}\selectfont
\begin{tabular}[t]{llll>{\raggedright\arraybackslash}p{10em}l}
\toprule
variable & dynamics & level & type & unit/comment & source\\
\midrule
Weekday & daily & national & categorical & Sat through Thu as six binary variables, Fri as baseline & -\\
Holiday (report) & daily & national & binary & - & -\\
Holiday (exposure) & daily & national & binary & - & -\\
\addlinespace[0.3em]
\multicolumn{6}{l}{\textbf{Mobility}}\\
\hspace{1em}retail and recreation & daily & state & numeric & percent change compared to reference period & Google \citep{google_mobility}\\
\hspace{1em}grocery and pharmacy & daily & state & numeric & percent change compared to reference period & Google \citep{google_mobility}\\
\hspace{1em}parks & daily & state & numeric & percent change compared to reference period & Google \citep{google_mobility}\\
\hspace{1em}workplaces & daily & state & numeric & percent change compared to reference period & Google \citep{google_mobility}\\
\hspace{1em}residential & daily & state & numeric & percent change compared to reference period & Google \citep{google_mobility}\\
\hspace{1em}transit stations & daily & state & numeric & percent change compared to reference period & Google \citep{google_mobility}\\
\addlinespace[0.3em]
\multicolumn{6}{l}{\textbf{Weather}}\\
\hspace{1em}Rainfall & daily & district & numeric & mm (l/sqm) & DWD \citep{dwd_weather}\\
\hspace{1em}Temperature & daily & district & numeric & °C & DWD \citep{dwd_weather}\\
\hspace{1em}Humidity & daily & district & numeric & relative humidity (\%) & DWD \citep{dwd_weather}\\
\hspace{1em}Wind & daily & district & numeric & m/s & DWD \citep{dwd_weather}\\
\addlinespace[0.3em]
\multicolumn{6}{l}{\textbf{Policies}}\\
\hspace{1em}Ban of mass gatherings & daily & national & binary & - & -\\
\hspace{1em}School and kindergarten closures & daily & state & numeric & 0 for no closure, 1 for full closure, 0.5 for partial reopening & -\\
\hspace{1em}Contact restrictions & daily & national & binary & - & -\\
\hspace{1em}Mandatory face masks & daily & district & binary & - & IZA \citep{mitze2020face}\\
\hspace{1em}Interventions & daily & district & numeric & sum of the above & -\\
\addlinespace[0.3em]
\multicolumn{6}{l}{\textbf{Socio-demographic}}\\
\hspace{1em}Age & constant & district & numeric & 2 variables: share of population >=65 years \& <18 years & INKAR \citep{inkar}\\
\hspace{1em}Gender & constant & district & numeric & share of female population & INKAR \citep{inkar}\\
\hspace{1em}Population density & constant & district & numeric & population per sqkm & INKAR \citep{inkar}\\
\hspace{1em}Foreign citizens & constant & district & numeric & 2 variables: share of foreign citizens \& of population seeking refuge & INKAR \citep{inkar}\\
\hspace{1em}Socio-economic status & constant & district & numeric & share of households with low income & INKAR \citep{inkar}\\
\hspace{1em}Turnout & constant & district & numeric & voter turnout in last election & INKAR \citep{inkar}\\
\hspace{1em}Right-wing populist party votes & constant & district & numeric & share of votes for AfD in last election & INKAR \citep{inkar}\\
\hspace{1em}Nursing homes & constant & district & numeric & number of nursing (retirement) homes & INKAR \citep{inkar}\\
\addlinespace[0.3em]
\multicolumn{6}{l}{\textbf{Awareness}}\\
\hspace{1em}Searches corona & daily & state & numeric & percent relative to other states and observation period & Google \citep{google_trends}\\
\hspace{1em}COVID-19 burden & daily & district & numeric & reported cases on day of exposure & RKI \citep{casenumbers_rki}\\
\addlinespace[0.3em]
\multicolumn{6}{l}{\textbf{Case numbers}}\\
\hspace{1em}Reported new cases of COVID-19 & daily & district & numeric & - & RKI \citep{casenumbers_rki}\\
\hspace{1em}Active cases & daily & district & numeric & active cases on day of report & RKI \citep{casenumbers_rki}\\
\bottomrule
\end{tabular}
\end{table}

All continuous variables but the outcome ``Reported new cases of COVID-19'' and the offset ``Active cases'' were centered and scaled by one standard deviation for numerical stability, while we left binary variables as-is. After estimating the effects of variables, we re-scaled continuous variables' effects to their original scale. Additionally for mobility, we re-transformed the effects of the principal components to the original mobility variables. Furthermore, we lagged the effect of all variables (but outcome, offset, and the non-dynamic socio-demographic variables) by 12 days (see Section \ref{results}) which means that we assumed that their effects on the outcome will be visible after 12 days.

@@

We used a directed acyclic graph as a graphical representation of the hypothesized causal reasoning that leads to exposure to the SARS-CoV-2 virus, onset of COVID-19, and finally reports of COVID-19 cases. We use the terms ``causal effect'' or ``causal relationship'' for effect estimates that are based on this causal graph framework. Every node \(v_i\) in the causal graph is the graphical representation of an observed or unobserved variable \(x_i\), a directed edge \(e_{ij}\) is an arrow from node \(v_i\) to \(v_j\) that implies a direct causal relationship from variable \(x_i\) onto variable \(x_j\). The set of all nodes is denoted by \(V\), the set of all edges by \(E\), as such, the complete DAG is the tuple \(G=(V,E)\). The seminal works of Spirtes and Pearl \citep{spirtes2000causation, pearl2009causality} introduce the theory of causal analysis, do-calculus, and how to analyze a DAG to estimate the total or direct causal effect from a variable \(x_i\) onto a variable \(x_j\). The direct effect is the effect associated with the edge \(e_{ij}\) only (if it exists), while the total effect takes indirect effects via other paths from \(v_i\) to \(v_j\) into account, too. Here we estimated total effects only, since most of our variables were not hypothesized to have a direct effect on the \emph{reported} number of new COVID-19 cases. In contrast to prediction tasks, where one would include all variables available, it is actually ill-advised to use all available variables to estimate causal effects, due to introducing bias by adjusting for unnecessary variables within the causal DAG. This is why we need to identify a valid set of necessary variables (an adjustment set) to estimate the proper causal effect \citep{pearl2009causality}. The ``minimal adjustment set'' \citep{greenland1999causal} is a valid adjustment set of variables that does not contain another valid adjustment set as a subset. However, identifying a minimal adjustment set might not be enough to reliably estimate the causal effect. Thus, we identified the ``optimal adjustment set'' \citep{henckel2019graphical} as the set of variables which is a valid adjustment set while having the lowest AIC.

We analyzed the DAG from Section \ref{causal-model} with the R Software \citep{rsoftware} and the R packages \texttt{dagitty} (formal representation of the graph and minimal adjustment sets \citep{textor_robust_2017}) and \texttt{pcalg} (for finding an optimal adjustment set \citep{pcalg}). For the defined exposures and the outcome ``Reported new cases of COVID-19'', we computed the minimal and optimal adjustment sets. Since it was possible that these sets contained unobserved variables that needed to be left out of the regression model, we chose the valid set with the lowest AIC (see next section) to estimate the final total causal effect from exposure to outcome.

@@

To reduce the effects of multicollinearity and to prevent overfitting, we used a ridge regression approach \citep{ridge1970}, which is a regularization method that shrinks regression coefficients to allow for better fits on unseen data and alleviates the effect of correlation between variables on their respective regression coefficients. The hyper-parameter \(\lambda\) of the ridge regression was chosen by 10-fold cross-validation, where the folds were constructed from subsets of the \(401\) districts.

We conducted a log-linear regression (function \texttt{glm} with \texttt{family=poisson()} for Poisson regression, and \texttt{glm.nb} from the \texttt{MASS} package for the negative binomial regression \citep{mass}) for the full data set to assess general model adequacy and to estimate the \(\theta\) parameter of the negative binomial. We used this parameter with the \texttt{cv.glmnet} function from the R package \texttt{glmnet} \citep{glmnet2010} with \texttt{family=negative.binomial(theta)} and chose the \(\lambda\) value within one standard deviation from the minimal \(\lambda\) as regularization hyper-parameter. Afterwards, we calculated the effects of separate exposures on the outcome. For every exposure, we analyzed the different valid adjustment sets given by analysis of the causal DAG (i.e.~the minimal and optimal adjustment sets). Then, we first checked if the respective set included unobserved variables. If this was the case for the optimal adjustment set, we discarded the unobserved variables from the set and checked if it was still a valid adjustment set (function \texttt{gac} in package \texttt{pcalg} \citep{perkovi2015complete}). If a minimal adjustment set contained unobserved variables, we discarded the whole set. If no valid adjustment set for a given exposure was available, we concluded that the effect of this exposure was unidentifiable within our causal graph. We used the function \texttt{glmnet} with the parameters \(\theta\) and \(\lambda\) as above on every remaining valid adjustment set as regressors (that is, we applied ridge regression) and calculated the Akaike information criterion (AIC) for this model/set of regressors. Finally, for every exposure, we decided for the model/adjustment set (if available) with the lowest AIC. We report the exponentiated estimated coefficients for the separate exposures on their original scale.

@@

\caption{\label{tab:causaleffects}Effect estimates from causal graph analysis}
\centering
\begin{tabular}[t]{lr}
\toprule
cause & effect estimate\\
\midrule
\addlinespace[0.3em]
\multicolumn{2}{l}{\textbf{Mobility}}\\
\hspace{1em}retail and recreation & 1.0010\\
\hspace{1em}grocery and pharmacy & 0.9988\\
\hspace{1em}parks & 0.9997\\
\hspace{1em}transit stations & 1.0024\\
\hspace{1em}workplaces & 1.0025\\
\hspace{1em}residential & 0.9914\\
\addlinespace[0.3em]
\multicolumn{2}{l}{\textbf{Awareness}}\\
\hspace{1em}Searches corona & 1.0007\\
\hspace{1em}COVID-19 burden & 0.9922\\
\addlinespace[0.3em]
\multicolumn{2}{l}{\textbf{Weather}}\\
\hspace{1em}Temperature & 0.9959\\
\hspace{1em}Rainfall & 1.0100\\
\hspace{1em}Humidity & 1.0042\\
\hspace{1em}Wind & 1.0164\\
\addlinespace[0.3em]
\multicolumn{2}{l}{\textbf{Policies}}\\
\hspace{1em}Interventions & 0.8037\\
\addlinespace[0.3em]
\multicolumn{2}{l}{\textbf{Demographic}}\\
\hspace{1em}Age (pop. 65 and older & 0.9976\\
\hspace{1em}Age (pop. younger 18 & 1.0097\\
\hspace{1em}Foreign citizens & 1.0034\\
\hspace{1em}Foreign citizens (refugees & 1.0010\\
\hspace{1em}Gender & 0.5232\\
\hspace{1em}Nursing homes & 1.0007\\
\hspace{1em}Population density & 1.0000\\
\hspace{1em}Socio-economic status & 0.9993\\
\bottomrule
\end{tabular}
\end{table}

We list the results of our causal analysis for the effects of different exposure variables in Table \ref{tab:causaleffects}. The estimates are multiplicative rates of increase/decrease for a one unit increase of the respective variable: Values above 1 lead to an increase, below 1 to a decrease of the infection rate. To put these estimates into perspective, Figure \ref{fig:effectsrange} shows the relative causal effect of the different exposure variables on the number of reported COVID-19 cases on a range of sensible values of the exposure variables (95 percent quantiles of data points).

@@

Within our framework, we saw significant effects for mobility in retail/recreational areas and essential shopping (grocery and pharmacy). For retail/recreation, an increase of 1 percent point mobility compared to the reference period (03 January to 06 February 2020) leads to an increase of the daily reported case number by about 0.1 percent. Contrarily, a corresponding increase of 1 percent point for the areas of grocery/pharmacy leads to a decrease in the reported case number by approximately 0.12 percent. Mobility on workplaces showed a small effect of 0.25 increase in case numbers for every 1 percent point increase in mobility. Other causal effects of mobility were insubstantial and not consistent in their direction (99 percent confidence intervals of estimates include 1). Figure \ref{fig:effectsrange} shows the effects of mobility on a range of possible values. Thus, we expect an increase of daily cases by approximately 3 percent if mobility in retail/recreation reaches baseline levels of 0 percent difference to the reference period. On the other hand, an increase of mobility for grocery/pharmacy by 10 percent points compared to the reference period leads to a reduction of the infection rate by approximately 2 percent.

``Awareness'' had two opposite effects on the outcome in our DAG. Awareness measured by Google searches for \emph{corona} had a positive effect on the number of reported cases. An one percent point increase of the state's Google searches (relative to other states and the observation period) leads to an increase of approximately 0.1 percent. For example, if a district shows 10 percent points more relative searches for \emph{corona} than another one, we expect approximately 1 percent more infections for this district after 8 days. \emph{COVID-19 burden} (reported number of cases on day of exposure) affected the outcome negatively, where every additional daily case in the district leads to a 0.8 percent decrease in newly reported case numbers. The last plot in Figure \ref{fig:effectsrange} visualizes this relationship: For a local outbreak with 20 daily cases as COVID-19 burden, we estimate as total causal effect a subsequent reduction of infection rate by 14.6 percent.

Within our model, we observed a causal effect of temperature and all other weather variables. Every increase of 1 degree Celsius in temperature leads to a reduction of the daily reported case numbers by approximately 0.4 percent. On the other hand, we found an increasing effect of rainfall: One millimeter (=1 liter per square meter) more rainfall leads to an increase of reported case numbers by approximately 1 percent. We observe effects for humidity and wind as well (higher humidity leading to more cases, stronger wind leading to less cases). In perspective (Figure \ref{fig:effectsrange}), with temperature we expect an increase by approximately 8.5 percent at a daily average temperature of \(0^{\circ}C\) compared to a day with \(20^{\circ}C\). For rainfall, we expect on a rainy day with 10 mm rainfall a corresponding increase of the infection rate by approximately 10.5 percent compared to a day with no precipitation.

The variable ``interventions'' showed \ldots{}

The effects of the different socio-demographic factors \ldots{}

In all cases we opted to use the (reduced) optimal adjustment set over the minimal adjustment sets because of lower AICs, except for exposure variable ``nursing homes'', where the minimal adjustment set had a lower AIC. We found that there were no valid adjustment sets for the unidentifiable variables turnout and right-wing populist party votes.

We decided for the lag of 12 days based on the lowest AIC value compared to other lags between 1 and 20 days on the full model. Similarly, negative binomial regression was chosen over Poisson regression, because the latter showed overdispersion and an higher AIC value.
